Forecasting climate change impacts on plant population dynamics at large spatial extents: a test case with sagebrush (\emph{Artemisia}) species

### 2015-11-13 docs/ms_sageAbundance.Rmd
edits to Andy's abstract edits

Abstract
--------
New remote sensing technology is producing ecological data at unprecedented spatial scales. These massive data offer an opportunity to study ecological processes at the landscape and regional scales that have traditionally been observed only at the level of small study plots. However, harnessing the power of new spatial data requires statistical models that can both rigorously account for spatial autocorrelation and can also be fit within a reasonable amount of computation time. Here we demonstrate how recent advances in the modeling of animal populations can be applied to newly available remotely-sensed species-specific estimates of plant cover at the landscape scale.  We fit a spatiotemporal population model to a 27 year time series of sagebrush (*Artemisia* spp.) percent cover over a $2.5\times5$ km landscape in southwestern Wyoming. We include interannual variation in precipitation and temperature as covariates in the model to investigtate how climate affects the cover of this species. (include more on the methodological advancement/novelty.. knots etc??) We then use the model to forecast the future abudance of sagebrush at the landscape scale under projected climate change. Our analysis provides spatially explicit estimates of increases and decreases in a plant population that have, until now, been impossible to produce at this scale. (any new insights about the statistical method?) Critically our large-scale and long-term predictions are rooted in small-scale and short-term population dynamics and so serve as a useful alternative to the predictions offered by most species distribution models that do not include population dynamics at all. We expect the approach we develope here will become an indispensable method for using new remote sensing data to model how populations respond to environmental changes, especially climate change, that play out at spatial scales far greater than the average study plot.

>>>>

Abstract
--------
New remote sensing technology is producing ecological data at unprecedented spatial scales. 
These massive data offer an opportunity to study ecological processes at the landscape and regional scales that have traditionally been observed only at the level of small study plots. 
However, harnessing the power of new spatial data requires statistical models that can both rigorously account for spatial autocorrelation and can also be fit within a reasonable amount of computation time. 
Here we demonstrate how recent advances in the modeling of animal populations can be applied to newly available remotely-sensed species-specific estimates of plant cover at the landscape scale.  
Using modern, computationally efficient model reparameterizations we fit a spatiotemporal population model to a 27 year time series of sagebrush (*Artemisia* spp.) percent cover over a $2.5\times5$ km landscape in southwestern Wyoming with rigorous accounting of spatial autocorrelation.
We include interannual variation in precipitation and temperature as covariates in the model to investigtate how climate affects the cover of this species.
We then use the model to forecast the future abudance of sagebrush at the landscape scale under projected climate change. 
Our analysis provides spatially explicit estimates of change in a plant population that have, until now, been impossible to produce at this scale. 
Critically our large-scale and long-term predictions are rooted in small-scale and short-term population dynamics and so serve as a useful alternative to the predictions offered by most species distribution models that do not include population dynamics at all. 
We expect the approach we develope here will become an indispensable method for using new remote sensing data to model how populations respond to environmental changes, especially climate change, that play out at spatial scales far greater than the average study plot.
